---
title: "Case Study 1"
author: "Seif Eldin Haybat"
date: "2022-11-10"
output: html_document
---


# How Does a Bike-Share Navigate Speedy Success?

## Scenario

You are a junior data analyst working in the marketing analyst team at Cyclistic, a bike-share company in Chicago. The director of marketing believes the company’s future success depends on maximizing the number of annual memberships. Therefore, your team wants to understand how casual riders and annual members use Cyclistic bikes differently. From these insights, your team will design a new marketing strategy to convert casual riders into annual members. But first, Cyclistic executives must approve your recommendations, so they must be backed up with compelling data insights and professional data visualizations.

## Characters and teams

● Cyclistic: A bike-share program that features more than 5,800 bicycles and 600 docking stations. Cyclistic sets itself apart by also offering reclining bikes, hand tricycles, and cargo bikes, making bike-share more inclusive to people with disabilities and riders who can’t use a standard two-wheeled bike. The majority of riders opt for traditional bikes; about 8% of riders use the assistive options. Cyclistic users are more likely to ride for leisure, but about 30% use them to commute to work each day.

● Lily Moreno: The director of marketing and your manager. Moreno is responsible for the development of campaigns and initiatives to promote the bike-share program. These may include email, social media, and other channels.

● Cyclistic marketing analytics team: A team of data analysts who are responsible for collecting, analyzing, and reporting data that helps guide Cyclistic marketing strategy. You joined this team six months ago and have been busy learning about Cyclistic’s mission and business goals — as well as how you, as a junior data analyst, can help Cyclistic achieve them.

● Cyclistic executive team: The notoriously detail-oriented executive team will decide whether to approve the recommended marketing program.

## Guiding questions

● What is the problem you are trying to solve?
The problem I am trying to solve is to build a profile for casual and annual members and how to convert casual riders into members.
● How can your insights drive business decisions?
My insights can drive business decisions by helping the marketing team come up with a marketing strategy to convert casual riders into annual members.

# Ask

Three questions will guide the future marketing program:

1.	How do annual members and casual riders use Cyclistic bikes differently?
2.	Why would casual riders buy Cyclistic annual memberships?
3.	How can Cyclistic use digital media to influence casual riders to become members?

## Key tasks

1.	Identify the business task
2.	Consider key stakeholders

## Deliverable

•	  A clear statement of the business task

# Prepare

I used Cyclistic’s historical trip data to analyze and identify trends. The data has been made available by Google here

## Guiding questions

● Where your data located?
My data is located in a html link.
● How is the data organized?
The data is separated by each month in it’s own .csv file.
● Are there issues with bias or credibility in this data? Does your data ROCCC?
There are no issues with bias or credibility in this data as it is gathered first hand by the company and the population is it’s entire client base. My data is reliable, organized, comprehensive, current and cited.
● How are you addressing licensing, privacy, security, and accessibility?
I am addressing privacy by not including credit card information in the data set and creating a unique ID for each rider. The company has their own license over the data
● How did you verify the data’s integrity?
Every dataset has columns that are labelled, and data is correctly populated under each type.
● How does it help you answer your question?
This will help answer my question because it can give keen insight into rider’s pattern of behavior.
● Are there any problems with the data?
There are some cells with empty or blank values.

## Key tasks

1.	Download data and store it appropriately.
2.	Identify how it’s organized.
3.	Sort and filter the data.
4.	Determine the credibility of the data.

## Deliverable

-[x] A description of all data sources used
The main data is from the last 12 months.

# Process

I will download the last 12 months worth of data and create some new columns labelled “ride_length” and, “day_of_the_week”.

## Guiding questions

● What tools are you choosing and why? 
I’m using Google Sheets and R. I’m using Google sheets to sort and organize my data because I’m comfortable with it and using R to present my findings.
● Have you ensured your data’s integrity? 
Yes, the data is consistent throughout the column.
● What steps have you taken to ensure that your data is clean? 
First, duplicated values were removed, then the columns were formatted to their correct data type.
● How can you verify that your data is clean and ready to analyze? 
It can be verified by this R markdown document,
● Have you documented your cleaning process so you can review and share those results? 
Yes, it is all documented in this R markdown file.

## Key tasks

1.	Check the data for errors.
2.	Choose your tools.
3.	Transform the data so you can work with it effectively.
4.	Document the cleaning process.

## Deliverable

Documentation of any cleaning or manipulation of data

# Analyze

The data exploration will consist of building a profile for annual members and how they differ from casual riders.
Putting in a new variable with a simpler name will help reduce some typing in the future.


## Set up
```{r}
library(tidyverse)  #helps wrangle data
library(lubridate)  #helps wrangle date attributes
library(ggplot2)  #helps visualize data
```

# STEP 1: COLLECT DATA
#======================================================
## Uploading datasets into R
```{r}
Q2_2019 <- read_csv("Divvy_Trips_2019_Q2.csv")
Q3_2019 <- read_csv("Divvy_Trips_2019_Q3.csv")
Q4_2019 <- read_csv("Divvy_Trips_2019_Q4.csv")
Q1_2020 <- read_csv("Divvy_Trips_2020_Q1.csv")
```

# STEP 2: WRANGLE DATA AND COMBINE INTO A SINGLE FILE
#======================================================
## Understand Data and Combine Datasets
### Look into Q1_2020 dataset column names to match other datasets to. While the names don't have to be in the same order, they DO need to match perfectly before we can use a command to join them into one file.

```{r}
colnames(Q2_2019)
colnames(Q3_2019)
colnames(Q4_2019)
colnames(Q1_2020)
```
## Rename columns  to make them consisent with q1_2020 (as this will be the supposed going-forward table design for Divvy)
```{r}
(Q4_2019 <- rename(Q4_2019
                   ,ride_id = trip_id
                   ,rideable_type = bikeid 
                   ,started_at = start_time  
                   ,ended_at = end_time  
                   ,start_station_name = from_station_name 
                   ,start_station_id = from_station_id 
                   ,end_station_name = to_station_name 
                   ,end_station_id = to_station_id 
                   ,member_casual = usertype))
(Q3_2019 <- rename(Q3_2019
                   ,ride_id = trip_id
                   ,rideable_type = bikeid 
                   ,started_at = start_time  
                   ,ended_at = end_time  
                   ,start_station_name = from_station_name 
                   ,start_station_id = from_station_id 
                   ,end_station_name = to_station_name 
                   ,end_station_id = to_station_id 
                   ,member_casual = usertype))
(Q2_2019 <- rename(Q2_2019
                   ,ride_id = "01 - Rental Details Rental ID"
                   ,rideable_type = "01 - Rental Details Bike ID" 
                   ,started_at = "01 - Rental Details Local Start Time"  
                   ,ended_at = "01 - Rental Details Local End Time"  
                   ,start_station_name = "03 - Rental Start Station Name" 
                   ,start_station_id = "03 - Rental Start Station ID"
                   ,end_station_name = "02 - Rental End Station Name" 
                   ,end_station_id = "02 - Rental End Station ID"
                   ,member_casual = "User Type"))
```

## Inspect the dataframes and look for incongruencies
```{r}
str(Q2_2019)
str(Q3_2019)
str(Q4_2019)
str(Q1_2020)
```

## Convert ride_id and rideable_type to character so that they can stack correctly
```{r}
Q4_2019 <-  mutate(Q4_2019, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type)) 
Q3_2019 <-  mutate(Q3_2019, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type)) 
Q2_2019 <-  mutate(Q2_2019, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type)) 
```

## Stack individual dataframes into one big dataframe and removing unused columns
```{r}
all_trips <- bind_rows(Q2_2019, Q3_2019, Q4_2019, Q1_2020)
all_trips <- all_trips %>%  
  select(-c(start_lat, start_lng, end_lat, end_lng, birthyear, gender, "01 - Rental Details Duration In Seconds Uncapped", "05 - Member Details Member Birthday Year", "Member Gender", "tripduration"))
```

# STEP 3: CLEAN UP AND ADD DATA TO PREPARE FOR ANALYSIS
#======================================================
## Inspect the new table that has been created
```{r}
colnames(all_trips)  #List of column names
nrow(all_trips)  #How many rows are in data frame?
dim(all_trips)  #Dimensions of the data frame?
head(all_trips)  #See the first 6 rows of data frame.  Also tail(all_trips)
str(all_trips)  #See list of columns and data types (numeric, character, etc)
summary(all_trips)  #Statistical summary of data. Mainly for numerics
```

## In the “member_casual” column, replace “Subscriber” with “member” and “Customer” with “casual”
#### Before 2020, Divvy used different labels for these two types of riders … we will want to make our dataframe consistent with their current nomenclature
```{r}
table(all_trips$member_casual) #To see how many observations fall under each usertype
all_trips <-  all_trips %>% 
  mutate(member_casual = recode(member_casual
                           ,"Subscriber" = "member"
                           ,"Customer" = "casual"))
table(all_trips$member_casual) #To validate that the proper number of observations was reassigned
```

## Add columns that list the date, month, day, and year of each ride.
#### This will allow us to aggregate ride data for each month, day, or year, before completing these operations we could only aggregate at the ride level
#### https://www.statmethods.net/input/dates.html more on date formats in R found at that link
```{r}
all_trips$date <- as.Date(all_trips$started_at) #The default format is yyyy-mm-dd
all_trips$month <- format(as.Date(all_trips$date), "%m")
all_trips$day <- format(as.Date(all_trips$date), "%d")
all_trips$year <- format(as.Date(all_trips$date), "%Y")
all_trips$day_of_week <- format(as.Date(all_trips$date), "%A")

```

## Add a "ride_length" calculation to all_trips (in seconds)
#### https://stat.ethz.ch/R-manual/R-devel/library/base/html/difftime.html
```{r}
all_trips$ride_length <- difftime(all_trips$ended_at,all_trips$started_at)
str(all_trips)
```

# Convert "ride_length" from Factor to numeric so we can run calculations on the data
```{r}
is.factor(all_trips$ride_length)
all_trips$ride_length <- as.numeric(as.character(all_trips$ride_length))
is.numeric(all_trips$ride_length)

```

## Remove "bad" data
#### The dataframe includes a few hundred entries when bikes were taken out of docks and checked for quality by Divvy or ride_length was negative
#### We will create a new version of the dataframe (v2) since data is being removed
#### https://www.datasciencemadesimple.com/delete-or-drop-rows-in-r-with-conditions-2/
```{r}
all_trips_v2 <- all_trips[!(all_trips$start_station_name == "HQ QR" | all_trips$ride_length<0),]
```

# STEP 4: CONDUCT DESCRIPTIVE ANALYSIS
#=====================================

## Descriptive analysis on ride_length (all figures in seconds)
```{r}
mean(all_trips_v2$ride_length) #straight average (total ride length / rides)
median(all_trips_v2$ride_length) #midpoint number in the ascending array of ride lengths
max(all_trips_v2$ride_length) #longest ride
min(all_trips_v2$ride_length) #shortest ride
summary(all_trips_v2$ride_length) #To have the four outputs from above in one line use summary() on the specific attribute
```

## Compare members and casual users
```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = mean)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = median)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = max)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = min)

```

## See the average ride time by each day for members vs casual users
```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
```

## Notice that the days of the week are out of order. Let's fix that and rerun it again.
```{r}
all_trips_v2$day_of_week <- ordered(all_trips_v2$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))

aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)

```

## analyze ridership data by type and weekday
```{r}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>%  #creates weekday field using wday()
  group_by(member_casual, weekday) %>%  #groups by usertype and weekday
  summarise(number_of_rides = n()				#calculates the number of rides and average duration 
            ,average_duration = mean(ride_length)) %>% 	# calculates the average duration
  arrange(member_casual, weekday)		# sorts

```

## Let's visualize the number of rides by rider type
```{r}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge")
```

## Let's create a visualization for average duration
```{r}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge")
```

# STEP 5: EXPORT SUMMARY FILE FOR FURTHER ANALYSIS
#=================================================
```{r}
Cyclistic <- aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)

write.csv(Cyclistic, "Cyclistic.csv")
```


## Guiding questions

● How should you organize your data to perform analysis on it?

Making sure all the columns are consistent as well as the data formatted within it.

● Has your data been properly formatted?

Yes, my data has been properly formatted.

● What surprises did you discover in the data?

I was surprised to find the data file template changed from 2019 to 2020.

● What trends or relationships did you find in the data?

I found that causal rider take longer rides and are more active on the weekend, Friday evening to Sunday evening. There are more members than casuals in the dataset, they take trips more often, for shorter durations and prefer docked bikes.

● How will these insights help answer your business questions?

These insights help us understand how riders use our product.


## Deliverable

A summary of your analysis

# Share

## Guiding questions

● Were you able to answer the question of how annual members and casual riders use Cyclistic bikes differently?

Yes I am.

● What story does your data tell?

Casual users and members use the Cyclistic service very differently but we can adjust to change casual to members.

● How do your findings relate to your original question?

My findings relate to my original question by building a profile for members, finding key differences between casual and members, and why each group uses bikes.

● Who is your audience? What is the best way to communicate with them?

My audience is the executive marketing team of Cyclistic.

● Can data visualization help you share your findings?

Data Visualizations can help me share my findings by summarizing all the complex data saved into a visual graphic. A picture (or graph) is worth a thousand words.

● Is your presentation accessible to your audience?

Yes, I worked to make sure my presentation is accessible to everyone.


## Deliverable

Supporting visualizations and key findings

# Act

## Guiding questions

● What is your final conclusion based on your analysis?

Members and casual riders have different riding profiles. Casual riders were found to take longer trips or rides on average than annual Members. This could mean Casual riders used Cyclistic bikes for leisure.
Casual riders often ride on weekends whereas annual Members use the program more over the week than on weekends. This could indicate that annual Members are using the bikes to commute to work.
Casual riders preferred using electric bikes to other ride types like classic and docked bikes. Annual Members used classic bikes more than other types. The reason Casual riders mostly rode electric bikes could be because they rode longer so, electric bikes eased propulsion.

● How could your team and business apply your insights?

My team and business could apply your insights by developing a marketing campaign to turn casual riders into members.

● What next steps would you or your stakeholders take based on your findings?

The next steps I would take would be to go deeper into the data to find more discoveries. At this point the marketing team could take this information to create their digital campaign.

● Is there additional data you could use to expand on your findings?

There is additional data I could use to expand my findings. The most recent monthly data will continue to develop and change, demonstrating to us the changing profile of members and casual riders as well as climate data to show trends with the weather.


## Deliverable

My recommendations based on my analysis

● I would recommend the marketing team to show the bikes as a good alternative to riding a car, showing the positive impact it has on the environment, getting exercise, and decreasing traffic congestion. 

● Highlight the availability of docking stations to show their reliable and readily available. 

● Users that take long rides could be offered a discount on renting bikes or better yet annual Members that travel more than a certain distance say for instance 10 miles should be offered a discount, also offering a coupon or discount during colder days to encourage rides.

● Cyclistic bike-share should introduce a weekend only membership that costs less than the current 7-day membership.

● A rebate program in form of refunds to annual Members that use a certain ride type more others should be introduced. It could be a small percentage of the cost.